Figure 3 - Motifs

### 2024-07-08 ms_figure_4_motifs.Rmd
update to new figure 4 (motifs)

# plot heatmap
# ... define heatmap font sizes
fs <- 5.5
fl <- 8

@@

# ... generate main heatmap with annotations
hm_3mers <- Heatmap(enr_3mers, name = "Pearson's resid.",
                    col = circlize::colorRamp2(
                        breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                        colors = colorRampPalette(enrichment_heatmap_colors)(64)),
                    cluster_rows = TRUE, show_row_dend = FALSE,
                    column_dend_reorder = match(colnames(enr_3mers),
                                                c("GCG", "CCG", "GGC", "GGG"),
                                                nomatch = 100),
                    cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
                    row_names_gp = gpar(fontsize = fl, fontfamily = "sans"),
                    column_names_gp = gpar(fontsize = fs, fontfamily = "mono"),
                    border = TRUE, border_gp = gpar(lwd = 0.5),
                    show_row_names = TRUE, show_column_names = TRUE,
                    use_raster = FALSE, show_heatmap_legend = TRUE,
                    bottom_annotation = kmer_annot,
                    heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                                labels_gp = gpar(fontsize = fl),
                                                border = "gray10",
                                                legend_direction = "horizontal",
                                                legend_width = unit(25, "mm"),
                                                title = "3-mer enrichment   ",
                                                title_position = "lefttop",
                                                title_gp = gpar(fontsize = fl)),
                    width = unit(130, "mm"), # heatmap body width
                    height = unit(3 * nrow(enr_3mers), "mm"))

@@

@@

# select motifs for detailed visualization
sel_motifs <- c(
    # known
    "Cdc10.m1",
    # novel
    "Toe2.m1", "Ams2.m1", "Esc1.m1",
    "SPAC3F10.12c.m1", "Zip1.m1", "Adn2.m4")
tf_names <- sub("[.]m[0-9]+$", "", sel_motifs)

# create motif visualizations
# ... setup plotting parameters
fgbg_col <- c(main_colors[1], main_colors[5]) # colors for foreground and background
gnm_col <- na_color # color for NA or "other"
w <- 2500  # width in bases around peak-center for meta-profiles
fs <- 9 # font size
nlabel <- 1 # number of k-mer words to label in enrichment plot

@@

    gg_ov_counts <- ggplot(pd_counts, aes(percentage, motif, fill = type)) +
        geom_col(position = position_stack()) +
        scale_x_continuous(labels = scales::label_percent(),
                           expand = expansion(mult = c(0, 0.15))) +
        scale_fill_manual(values = structure(c(gnm_col, fgbg_col[2:1]),
                                             names = levels(pd_counts$type))) +
        geom_text(data = pd_counts |> filter(grepl("foreground", type)),
                  aes(x = 1, label = sprintf("%.1f%%", percentage * 100)),
                  colour = fgbg_col[1], hjust = -0.1, vjust = 0.5,
                  size = fs / ggplot2::.pt,
                  fontface = "bold") +
        labs(x = "Fraction of motif sites",
             y = element_blank(),
             fill = "Genomic region: ") +
        theme_cowplot(fs) +
        theme(legend.position = "none") +
        guides(fill = guide_legend(nrow = 3))

@@

```{r Fig3}
#| fig.width: 11
#| fig.height: 15.5

cowplot::plot_grid(
    cowplot::plot_grid(
        cowplot::plot_grid(ZnII2Cys6_kmer_1, scale = 0.85),
        hm_ZnII2Cys6_sel_3mers,
        labels = c("A", "B"),
        rel_widths = c(2, 3.2),
        nrow = 1
    ),
    cowplot::plot_grid(
        plotlist = motif_characterization_plots,
        ncol = 1,
        rel_heights = c(rep(1, length(motif_characterization_plots) - 1), 0.4)),
    nrow = 2,
    labels = c("", "C"),
    align = "vh",
    axis = "b",
    rel_heights = c(1.4, 3.7)
)
```

>>>>

# plot heatmap
# ... define heatmap font sizes
fs <- 5.5 # small
fl <- 8   # large
ft <- 10  # title

@@

# ... generate main heatmap with annotations
hm_3mers <- Heatmap(enr_3mers, name = "Pearson's resid.",
                    col = circlize::colorRamp2(
                        breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                        colors = colorRampPalette(enrichment_heatmap_colors)(64)),
                    cluster_rows = TRUE, show_row_dend = FALSE,
                    column_dend_reorder = match(colnames(enr_3mers),
                                                c("GCG", "CCG", "GGC", "GGG"),
                                                nomatch = 100),
                    cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
                    row_names_gp = gpar(fontsize = fl, fontfamily = "sans"),
                    column_names_gp = gpar(fontsize = fs, fontfamily = "mono"),
                    column_title = "Binuclear zinc cluster TFs",
                    column_title_side = "top", column_title_gp = gpar(fontsize = ft),
                    border = TRUE, border_gp = gpar(lwd = 0.5),
                    show_row_names = TRUE, show_column_names = TRUE,
                    use_raster = FALSE, show_heatmap_legend = TRUE,
                    bottom_annotation = kmer_annot,
                    heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                                labels_gp = gpar(fontsize = fl),
                                                border = "gray10",
                                                legend_direction = "horizontal",
                                                legend_width = unit(25, "mm"),
                                                title = "3-mer enrichment   ",
                                                title_position = "lefttop",
                                                title_gp = gpar(fontsize = fl)),
                    width = unit(130, "mm"), # heatmap body width
                    height = unit(3 * nrow(enr_3mers), "mm"))

@@

```{r peak-overlaps-within-tf-family}
#| fig.width: 10
#| fig.height: 8

# get enrichment calls for specific peaks that are
# bound by at least one of the binuclear zinc cluster TFs used above
pd <- as.data.frame(is_enr_specific[, tf_names]) |>
    tibble::rownames_to_column("peak_name") |>
    pivot_longer(cols = all_of(tf_names),
                 values_to = "enriched",
                 names_to = "tf_name") |>
    filter(enriched) |>
    group_by(peak_name) |>
    summarise(tfs_enr = list(tf_name))

# create upset plot
upset_ZnII2Cys6_sel <- ggplot(pd, aes(x = tfs_enr)) +
        geom_bar() +
        scale_x_upset() +
        labs(x = element_blank(),
             y = "Number of enriched peaks") +
        theme_cowplot()
upset_ZnII2Cys6_sel
```

@@

# select motifs for detailed visualization
sel_motifs <- c( "Toe2.m1", "Esc1.m1", "SPAC3F10.12c.m1",
                 "Ams2.m1", "Zip1.m1", "Adn2.m4")
tf_names <- sub("[.]m[0-9]+$", "", sel_motifs)

# create motif visualizations
# ... setup plotting parameters
fgbg_col <- c(main_colors[1], main_colors[5]) # colors for foreground and background
gnm_col <- bg_color # color for NA or "other"
w <- 2500  # width in bases around peak-center for meta-profiles
fs <- 9 # font size
nlabel <- 1 # number of k-mer words to label in enrichment plot

@@

    gg_ov_counts <- ggplot(pd_counts, aes(percentage, motif, fill = type)) +
        geom_col(position = position_stack()) +
        scale_x_continuous(labels = scales::label_percent(),
                           expand = expansion(mult = c(0, 0.15))) +
        scale_fill_manual(values = structure(c(gnm_col, fgbg_col[2:1]),
                                             names = levels(pd_counts$type))) +
        geom_text(data = pd_counts |> filter(grepl("foreground", type)),
                  aes(x = 1, label = sprintf("%.1f%%", percentage * 100)),
                  colour = fgbg_col[1], hjust = -0.1, vjust = 0.5,
                  size = fs / ggplot2::.pt,
                  fontface = "bold") +
        labs(x = "Percent of motif sites",
             y = element_blank(),
             fill = "Genomic region: ") +
        theme_cowplot(fs) +
        theme(legend.position = "none") +
        guides(fill = guide_legend(nrow = 3))

@@

```{r Fig3}
#| fig.width: 11
#| fig.height: 14

cowplot::plot_grid(
    cowplot::plot_grid(
        cowplot::plot_grid(ZnII2Cys6_kmer_1, scale = 0.85),
        hm_ZnII2Cys6_sel_3mers,
        labels = c("A", "B"),
        rel_widths = c(2, 3.2),
        nrow = 1
    ),
    cowplot::plot_grid(
        plotlist = motif_characterization_plots,
        ncol = 1,
        rel_heights = c(rep(1, length(motif_characterization_plots) - 1), 0.4)),
    nrow = 2,
    labels = c("", "C"),
    align = "vh",
    axis = "b",
    rel_heights = c(1.4, 3.2)
)
```
